Binarity's Biases on Transit Survey Occurrence Rates

### 2018-02-23 paper/main.tex
jnw understanding errors edits

Imagine a group of astronomers that wants to measure the mean number of
planets of a certain type per star of a certain type.  They perform a
wide-field photometric search for planets that transit, and discover
all for which
\begin{equation}
  \frac{\delta}{\sigma}
  >
  \left(\frac{{\rm S}}{{\rm N}}\right)_{\rm min}.
\label{eq:S_N_thresh}
\end{equation}
Here the signal S is the observed transit depth $\delta$, in units of
relative flux.  The noise N is the uncertainty in the determination of
the flux ratio of the star inside and outside of transits.  The
minimum signal-to-noise for detection, $({\rm S}/{\rm N})_{\rm min}$,
depends on the properties of the telescope.  We will assume throughout
this paper that the photometric uncertainty scales as the inverse
square root of the number of photons collected from the source, {\it
i.e.}, the dependence on stellar properties is $\sigma \propto (L_{\rm
tot}/d^2)^{-1/2}$, where $L_{\rm tot}$ is the total luminosity of the
stellar system and $d$ is its distance from Earth.  The observed
signal amplitude $\delta$ depends on the fraction of the total
luminosity that comes from the planet-hosting star, since light from
unresolved stellar companions is not affected by the planet's transit.
% The transit duration also affects detectability, but we omit it in
% this work for brevity.

The astronomers analyze their data assuming that all stars are
single.  In particular they do not have accurate enough parallaxes to
tell that some of the stars are apparently overluminous.  They want to
compute the number of planets with size $r$ per star of mass $M$.  If
the total number of stars for which such planets could be detected
($N_\star$) and the number of detected planets ($N_{\rm det}$) are
both large, then the astronomers' estimate for the occurrence rate
($\Lambda$) is simply
\begin{equation}
  \Lambda = \frac{N_{\rm det}}{N_\star}
                    \, \frac{1}{p_{\rm tra}},
\label{eq:occ_rate_simple}
\end{equation}
where the geometric transit probability, $p_{\rm tra}$, accounts for
the fact that most planetary orbits are not aligned close enough with
our line of sight to produce transits.

There are many potential pitfalls in this calculation.  Some genuine
transit signals are missed even if they formally exceed the
signal-to-noise threshold, because of the probabilistic nature of
transit detection.  Planets can be misclassified due to statistical
and systematic errors in the catalogued properties of the stars.  Some
apparent transit signals are spurious, arising from noise fluctuations
or failures of ``detrending'' the astrophysical or instrumental
variations in the photometric signal.  Poor angular resolution leads
to blends between eclipsing binary stars and other stars along nearly
the same line of sight, producing signals that mimic those of
transiting planets.

We will focus exclusively on problems that arise from the fact that
many stars exist in gravitationally bound binary systems.  Even with
this narrow focus, there are numerous sources of error.  All three of
the quantities in Equation~\ref{eq:occ_rate_simple} are subject to
observational bias:
\begin{enumerate}
%    
    \item The detected planet count, $N_{\det}$, is actually the
      number of detected planets that {\it appear} to have size $r$,
      orbiting stars that {\it appear} to have mass $M$.  Whenever the
      planet-hosting star is part of a binary,
%    
    \begin{itemize}
        \item the planet's size will be misclassified because of the
          reduction in the amplitude of the photometric signal;
%        
        \item the host star's properties could be misclassified
          because its light is combined with a second star of a
          different spectral type.
%        
    \end{itemize}
%    
    \item The apparent number of stars that were searched, $N_\star$,
      is biased
%    
    \begin{itemize}
%        
        \item toward lower values, because it does not include all of
          the secondary stars that were searched for transiting
          planets;
%        
        \item toward higher values, because some of the stars that
          appeared to be suitably small and bright to detect the
          desired type of planet are in fact binaries for which the
          amplitude of the photometric signal would be reduced to an
          undetectable level.
%        
    \end{itemize}
%    
    \item The transit probability $p_{\rm tra}$, which scales with the
      stellar mean density as $\rho^{-1/3}$, is biased because the
      planet-hosting star could be misclassified.
%    
\end{enumerate}

Two other problems that may arise are not represented in
Equation~\ref{eq:occ_rate_simple}.  One is astrophysical: the true
occurrence rate of the desired type of planet may depend on whether
the host is a single star, the primary star of a binary, or the
secondary star of a binary.  Such differences could be caused by the
requirement for long-term dynamical stability, or differences in the
planet formation process.  When the search sample includes both
singles and binaries, the detected planets are thereby drawn from
different occurrence 
distributions~\citep[see][]{
    wang_occurrence_2015,kraus_impact_2016}.

The other problem is an observational effect. In the searched sample,
the ratio between the number of binary and single stars will differ
from that in a volume-limited sample due to a type of Malmquist bias.
The total luminosity of a binary is larger than the luminosity of
either the primary or secondary star.  This means that for transit
signals of a given amplitude, the binaries are searchable at greater
distances from the Earth.  Thus for any planet and star pair, adding a
stellar companion both helps and hinders transit detection: the extra
photons help by lowering the noise, and hinder by diminishing the
observed transit depth.

Planet occurrence rates are often presented as two-dimensional density
functions in the space of planet radius and orbital period.
Thankfully none of the errors we have enumerated lead to errors in the
period distribution.  When more than one transit is detected (as is
usually required by the surveyors), the orbital periods can be
measured without ambiguity regardless of whether the host star is
single or one member of a binary.  Hence in what follows, we focus on
the radius distribution and assume the periods are measured without
significant uncertainty.

Given all of the confusing and opposing sources of error, we will
proceed in stages. We start from a barebones model in which everything
can be written down on the back of a napkin, and build up to an
analytic model allowing for generality in the distribution of the
binaries and the planets they host.

>>>>

Imagine that a group of astronomers wants to measure the mean number
of planets per star.  They are particularly interested in planets of
radius $r$ and stars of mass $M$ and radius $R$.  They obtain a time
series of images of some region of the sky, and prepare light curves
for a large number of unresolved sources.  Then they search these
light curves for transit signals and detect all the signals for which
\begin{equation}
  \frac{\delta}{\sigma}
  >
  \left(\frac{{\rm S}}{{\rm N}}\right)_{\rm min}.
\label{eq:S_N_thresh}
\end{equation}
Here the signal, $\delta$, is the observed transit depth, the
dimensionless fraction by which the total light fades during transits.
Note that although $\delta$ is often equated with $(r/R)^2$, this is
not true when the host star is a member of an unresolved binary.  In
those cases, $\delta$ is smaller than $(r/R)^2$ because of the
constant light from the binary companion, an effect often called the
``dilution'' of the transit signal.  The noise, $\sigma$, is the
fractional uncertainty in the determination of the flux of the source,
which may include multiple stars that are blended together.  The
threshold signal-to-noise ratio depends on the desired level of
confidence that the signal is real.
% The transit duration also affects detectability, but we omit it in
% this work for brevity.

The astronomers analyze their data assuming that all the sources are
single stars.  In particular they do not have accurate enough
parallaxes to tell that some of the stars appear to be overluminous.
They count the number $N_{\rm det}$ of transit signals that appear to
be produced by the desired type of planet around the desired type of
star.  They also count the number $N_\star$ of ``searchable stars'' in
their survey, i.e., the number of stars of the desired type that are
bright enough to have allowed for the detection of a transit signal
with amplitude $(r/R)^2$.  They estimate the occurrence rate to be
\begin{equation}
  \Lambda = \frac{N_{\rm det}}{N_\star}
                    \, \frac{1}{p_{\rm tra}},
  \label{eq:occ_rate_simple}
\end{equation}
where the geometric transit probability, $p_{\rm tra}$, accounts for
the fact that most planetary orbits are not aligned close enough with
our line of sight to produce transits.

There are many potential pitfalls in this calculation.  Some genuine
transit signals are missed even if they formally exceed the
signal-to-noise threshold, because of the probabilistic nature of
transit detection.  Planets can be misclassified due to statistical
and systematic errors in the catalogued properties of the stars.  Some
transit-like signals are spurious, arising from noise fluctuations or
failures of ``detrending'' the astrophysical or instrumental
variations in the photometric signal.  Poor angular resolution leads
to blends between eclipsing binary stars and other stars along nearly
the same line of sight, producing signals that mimic those of
transiting planets.

Here, though, we will focus exclusively on problems that arise from
the fact that many stars exist in gravitationally bound binary
systems.  We will also focus on the errors in planet occurrence as a
function of radius, rather than orbital period. This is because when
more than one transit is detected (as is usually required by the
surveyors), the orbital periods can be measured without ambiguity
regardless of whether the host star is single or one member of a
binary.  Even with this narrow focus, there are numerous sources of
error.  All three of the quantities in
Equation~\ref{eq:occ_rate_simple} are biased:
\begin{enumerate}
%    
    \item The number of detected planets, $N_{\det}$, is actually the
      number of detected planets that {\it appear} to have size $r$,
      orbiting stars that {\it appear} to have mass $M$.  Whenever the
      planet-hosting star is part of a binary,
%    
    \begin{itemize}
        \item the planet's size could be misclassified because of the
          reduction in the amplitude of the photometric signal;
%        
        \item the host star's properties could be misclassified
          because its light is combined with a second star of a
          different spectral type.
%        
    \end{itemize}
%    
    \item The number of searchable stars, $N_\star$, is biased
%    
    \begin{itemize}
%        
        \item toward lower values, because it does not include all of
          the secondary stars that were inadvertently searched for
          transiting planets;
%        
        \item toward higher values, because some of the stars that
          appeared to be searchable are in fact binaries for which the
          amplitude of the photometric signal would have been reduced
          to an undetectable level.
%        
    \end{itemize}
%    
    \item The transit probability $p_{\rm tra}$ is biased because the
      planet-hosting star could be misclassified.  At fixed orbital
      period, the transit probability is proportional to
      $\rho^{-1/3}$, where $\rho$ is the stellar mean
      density~\citep{Winn2010}.  Therefore, any errors in determining
      the host star's mean density lead to errors in the correction
      for the transit probability.
%    
\end{enumerate}

There are at least two other complications that may arise, which are
not represented in Equation~\ref{eq:occ_rate_simple}.  The first one
is an observational effect. Within the sample of apparently searchable
stars, the ratio between the number of binary and single stars will
differ from the ratio that would be found in a volume-limited sample.
This is due to a type of Malmquist bias.  The total luminosity of a
binary is larger than the luminosity of either the primary or
secondary star.  This means that for transit signals of a given
amplitude, sources that are binaries appear to be searchable at
greater distances from the Earth.  Binaries are therefore
over-represented in the collection of apparently searchable stars.

The other complication is astrophysical: the true occurrence rate of a
certain type of planet may depend on whether the host is a single
star, the primary star of a binary, or the secondary star of a binary.
The rate might also depend on the characteristics of the binary, such
as the mass ratio and orbital period.  Such differences could be
caused by the requirement for long-term dynamical stability, or
differences in the planet formation process.  When the search sample
includes both singles and binaries, the detected planets are thereby
drawn from different occurrence distributions~\citep[see][]{
  wang_occurrence_2015,kraus_impact_2016}.

Given all of the confusing and opposing sources of error, we will
proceed in stages.  We start with a model so simple that everything
can be written down on the back of a napkin, and build up to an
analytic model allowing for generality in the distribution of the
binaries and the planets they host.
